Case Study: RNA-Seq Differential Analysis

### 2018-03-15 datasets/RNA-Seq/RNA-Seq.Rmd
Added 2nd dataset.

---
title: "Case Study: RNA-Seq Differential Analysis"
author: "Chinmay Shukla, Patrick Kimes"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
    html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

@@

# Data Preparation

@@

# Data Analysis

## Differential Testing

@@

## Covariate Diagnostics

@@

### Mean Coverage

@@

## Multiple-Testing Correction

@@

## Benchmark Metrics

@@

>>>>

---
title: "Case Study: RNA-Seq Differential Analysis"
author: "Chinmay Shukla, Patrick Kimes"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
    html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
bibliography: bibliography.bib
---

@@

# Dataset 1

## Data Preparation

@@

## Data Analysis

### Differential Testing

@@

### Covariate Diagnostics

@@

### Mean Counts

@@

### Multiple-Testing Correction

@@

### Benchmark Metrics

@@

# Dataset 2

As a second dataset, we will test for differences in gene expression upon the knockout of the microRNA mir-200c [@Kim_2013].  The raw fastq files can be found under the accession number `SRP030475`. As the number of samples is limited the experiment might be underpowered, as in most RNA-seq analysis. This is an experimental scenario that could benefit from power gaining from using modern FDR control methods.

## Data Preparation

We will download the pre-processed gene level counts available through recount2.

```{r}
download_study('SRP030475')
load(file.path('SRP030475', 'rse_gene.Rdata'))
dsd <- scale_counts(rse_gene)
```

We then subset for samples containing the control samples and the samples where mir200c was knocked down. Recount2 downloads data as a RangeSummarizedExperiment object, so we convert this into a DESeqDataSet object.

```{r}
dsd <- dsd[,grepl("WT|200c", colData(dsd)$title)]
colData(dsd)$mir200c <- factor(ifelse( grepl( "WT", colData(dsd)$title), "WT", "KO"))
dsd <- as( dsd, "DESeqDataSet" )
storage.mode(assays(dsd)[["counts"]]) <- "integer"
```

## Data Analysis

### Differential testing

Then, we set the design parameter to test for differences in expression upon mir200c knockout and run DESeq2. Similarly to the previous dataset, we set the parameter `independentFiltering=FALSE`.

```{r}
design(dsd) <- ~ mir200c
dsd <- DESeq(dsd)
res <- results(dsd, independentFiltering=FALSE) %>% 
  as.data.frame(  ) %>%
  na.omit() %>% 
  select( pvalue, baseMean, log2FoldChange, lfcSE, stat ) %>%
  rename( pval=pvalue, ind_covariate=baseMean, 
          effect_size=log2FoldChange, SE=lfcSE, 
          test_statistic=stat ) 
head( res )
```

### Mean Counts

As with the previous example, the mean counts is an informative covariate.

```{r, meancov-diag-scatter, fig.width=4.5, fig.height=3.5}
rank_scatter( res, pvalue = "pval", covariate = "ind_covariate") +
    ggtitle("Mean coverage as independent covariate") +
    xlab("Mean Expression")
```

Similarly to the first dataset, keeping all the tests results in a strange discretness. This is removed once we filter very lowly expressed genes.

```{r, meancov-diag-hist, fig.width=10, fig.height=3.2}
strat_hist( res, pvalue = "pval",
           covariate = "ind_covariate", maxy = 7.5)
res <- filter(res, ind_covariate > 1 )
strat_hist( res, pvalue = "pval",
           covariate = "ind_covariate", maxy = 3)
```

### Multiple-Testing Correction

```{r}
bd <- initializeBenchDesign()
sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
```

### Benchmark Metrics

```{r}
rejections_scatter(sb, supplementary = FALSE)
rejection_scatter_bins(sb, covariate = "ind_covariate",
                       bins = 4, supplementary = FALSE)
```

```{r}
plotFDRMethodsOverlap(sb, alpha = 0.05, nsets = ncol(sb),
                      order.by = "freq", decreasing = TRUE,
                      supplementary = FALSE)
```

```{r}
methods <- c("ashs", "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf", "scott-empirical")
plotCovariateBoxplots(sb, alpha = 0.1, nsets = 6,
                      methods = methods, trans = log2, maxNum = 5)
```
